---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r echo=TRUE}
# set environment
options(scipen=999)
base_dir <- "~/Documents/coursera/RepRes_2018/RepData_PeerAssessment1"
library(ggplot2)
library(dplyr)

# get data
setwd(base_dir)
unzip("activity.zip")
activity_df0 <- read.table(file="activity.csv", sep=",", header=TRUE)
```




## What is mean total number of steps taken per day?
```{r echo=TRUE}
# remove rows with NA values
no_na_df <- activity_df0[!is.na(activity_df0$steps), ]

# calculate total steps by date
steps_per_day_df <- aggregate(steps ~ date, no_na_df, sum)

# histogram of steps per day
ggplot(data=steps_per_day_df, aes(steps_per_day_df$steps)) + 
  geom_histogram() + 
  labs(title="Histogram of Total Steps Per Day (disregarding NA values)") +
  labs(x="Steps Per Day", y="Count")

# avg steps per day
no_na_mean = mean(steps_per_day_df$steps)
no_na_median = median(steps_per_day_df$steps)
no_na_mean
no_na_median
```

After discounting rows with missing values of the steps column, the mean number of steps per day was `r round(no_na_mean)`.  The median number of steps per day was `r round(no_na_median)`.

## What is the average daily activity pattern?
```{r echo=TRUE}

# plot avg steps over time
steps_per_interval_df <- aggregate(steps ~ interval, no_na_df, mean)
ggplot(data=steps_per_interval_df, aes(x=interval, y=steps, group=1)) +
  geom_line() +
  geom_point() +
  labs(title="Avg Steps Per Interval (disregarding NA values)") +
  labs(x="Interval", y="Steps")

# find interval with max average steps
max_interval = steps_per_interval_df %>% top_n(1, steps)
max_interval
```

The interval with the highest mean number of steps was `r round(max_interval[1])` with and average of `r round(max_interval[2])` steps.


## Imputing missing values
```{r echo=TRUE}

# count rows with steps=NA
na_rows = nrow(activity_df0[is.na(activity_df0$steps), ])
na_rows

# replace NA values with mean for respective interval (dplyr)
fill_na_df <- activity_df0 %>% 
  rowwise() %>%
  mutate(steps = if_else(is.na(steps), as.integer(round(steps_per_interval_df[ which(steps_per_interval_df$interval==interval), ]$steps)), steps))

# calculate total steps by date
steps_per_day_filled_df <- aggregate(steps ~ date, fill_na_df, sum)

# histogram of steps per day
ggplot(data=steps_per_day_filled_df, aes(steps_per_day_filled_df$steps)) + 
  geom_histogram() + 
  labs(title="Histogram of Total Steps Per Day (imputing NA values)") +
  labs(x="Steps Per Day", y="Count")

# avg steps per day
filled_mean = mean(steps_per_day_filled_df$steps)
filled_median = median(steps_per_day_filled_df$steps)
filled_mean
filled_median
```

The original data set contained `r na_rows` rows for which the steps column was missing a value.  For each of these rows, the steps value was imputed using the mean value for the associated time interval across the entire data set.

After imputing missing values for the steps column, the mean number of steps per day was `r round(filled_mean, 1)`.  The median number of steps per day was `r round(filled_median, 1)`.


## Are there differences in activity patterns between weekdays and weekends?
```{r}

# create factor var to label each row as weekday / weekend
fill_na_df <- fill_na_df %>% mutate(weekday = ifelse(weekdays(as.Date(date)) %in% c("Saturday", "Sunday"), "weekend", "weekday"))
fill_na_df$weekday <- factor(fill_na_df$weekday)

# plot avg steps over time, by weekday / weekend
steps_per_interval_df2 <- aggregate(steps ~ interval+weekday, fill_na_df, mean)
ggplot(data=steps_per_interval_df2, aes(x=interval, y=steps, group=1)) +
  geom_line() +
  geom_point() +
  labs(title="Avg Steps Per Interval") +
  labs(x="Interval", y="Steps") + 
  facet_grid(weekday ~ .)

```

